Schema Evolution in Interactive Programming Systems

### 2024-08-23 paper.tex
revise section 2: Layers and Dimensions of Schema Evolution

\section{Type Evolution}
To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
(except for monuments) evolve anyway, however poorly, because the usages in and around them are
changing constantly.'' Programs evolve both while they are being created and during their operation
in response to new needs. As with buildings, different aspects of program evolve at different pace
(Figure~\ref{fig:layers}). And ``because of the different rates of change of its components,
a program is always tearing itself apart.''

\subsection{Types, Data and Code}
In both buildings and programs, making a change at surface layers (skin or stuff) is relatively easy,
but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
In the case of programs, we refer to the more fundamental layer as \emph{types}, although we do not
use the term in a precise technical sense. Types is what defines the shape of data and code.
Types may be database schema, data types that define the shape of data, they may be explicitly
declared in code and checked statically, as well as dynamically checked or even implicit.
The key property of types for us is that they constrain the shape of other layers
of a program, namely data and code. As a result, a change of program types requires a
corresponding change in program data and code.

@@

The situation is illustrated on the right. We start with code $C_1$ and
data $D_1$ that have a shape corresponding to type $T_1$. When the type changes from $T_1$ to $T_2$ ($\rightarrow$),
a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
of \emph{type evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
to the new type $T_2$.
%
The complexity of the problem varies and finding a suitable transformation may require user
input. If there is code, it typically needs to be transformed, but data can sometimes be
discarded. We refer to such transformations done by a single user as \emph{local} and draw
them inside a box.

@@

Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
variants that exist independently and changes made to them may need to be synchronized. The
situation is illustrated on the right. Here, the initial type $T_1$ is transformed by two
users independently into types $T_2$ and $T_3$. The corresponding data is transformed accordingly,
producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
obtaining a type $T_4$ and data $D_4$ that adopt both of the changes.

@@

\subsection{Conceptual Framework}

We now describe a conceptual framework that lets us view a diverse set of real-world
type evolution problems in a unified way. We distinguish between two dimensions.
The \emph{local dimension} is concerned with how type evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how type evolution
propagates across different program variants that may exist independently.

\paragraph{Program Layers.}
We consider the more permanent layer of \emph{types} and two less permanent layers of code and
data that are structured by types. In specific case studies, the layers may map to
different aspects of programs, some may be not present and some may be implicit.

\begin{itemize}
\item \emph{Types} -- structure program data and determine aspects of code. A change in types
  requires a corresponding change in data and code. Types may be explicit (database schema, type definition)
  or implicit (shape of data structures, shape of a document).
\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
  database) or less permanent (state in live programming). When types evolve, data needs to be
  updated or (in case of transient state) discarded.
\item \emph{Code} -- or program logic, but not including source code that defines types.
  When types evolve, code needs to change to consume/produce data of a correct shape.
\end{itemize}

\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their types, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
type evolution in future programming systems. We consider two characteristics.

\begin{itemize}
\item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
  eventually adopt all changes. It may not be possible to adopt a change without adopting
  all earlier changes. In the divergence model, a user may choose only
  particular changes they want to adopt, but keep other aspects of their custom design.
\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (type
  evolution) can only originate from a particular source. In the decentralized model,
  changes can be done on any of the multiple co-existing variants of a program.
\end{itemize}

\paragraph{Challenge Problems.}
The conceptual framework offers a wide range of configurations for challenge problems, but
our aim is not to enumerate all options. We use it to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
type evolution. The case studies show how well-known solutions or problems in one domain
map to another domain and they also provide and inspiration and a benchmark for
developers of future programming systems.

>>>>

\section{Layers and Dimensions of Schema Evolution}

To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
(except for monuments) evolve anyway, however poorly, because the usages in and around them are
changing constantly.'' Programs evolve both while they are being created and during their operation
in response to new needs. As with buildings, different aspects of programs evolve at different paces
(Figure~\ref{fig:layers}). And ``because of the different rates of change of its components,
a program is always tearing itself apart.''

%\subsection{Schema, Data and Code}

In both buildings and programs, making a change at surface layers (skin or stuff) is relatively easy,
but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
In the case of programs, we refer to the more fundamental layer as \emph{schema}, although we do not
use the term in a precise technical sense. Schema are what define the shape of data and code, either as database schema, data types declared explicitly in code and checked statically, dynamically checked types and specifications, and even expectations implicit in the code.
The key property of schema for us is that they constrain the shape of other layers
of a program, namely data and code. As a result, a change of schema requires a
corresponding change in data and code.

@@

The situation is illustrated on the right. We start with code $C_1$ and
data $D_1$ that have a shape corresponding to schema $T_1$. When the schema changes from $T_1$ to $T_2$ ($\rightarrow$),
a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
of \emph{schema evolution} is synthesising ($\Rightarrow$) suitable corresponding transformations
that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
to the new schema $T_2$.
%
The complexity of the problem varies and finding a suitable transformation may require user
input. If there is code, it typically needs to be transformed, but data can sometimes be
discarded. We refer to such transformations done by a single user as \emph{local} and draw
them inside a box.

@@

Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
variants that exist independently and changes made to them may need to be synchronized. The
situation is illustrated on the right. Here, the initial schema $T_1$ is transformed by two
users independently into schema $T_2$ and $T_3$. The corresponding data is transformed accordingly,
producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
obtaining a schema $T_4$ and data $D_4$ that adopt both of the changes.

@@

%\subsection{Conceptual Framework}

\paragraph{Locality of Change.}
We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
The \emph{local dimension} is concerned with how schema evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how schema evolution
propagates across different program variants that may exist independently.

\paragraph{Program Layers.}
We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
data that are structured by those schema. In specific case studies, the layers may map to
different aspects of programs, some may be not present, and some may be implicit.

\begin{itemize}
\item \emph{Schema} -- structure program data and determine aspects of code. A change in schema
  requires a corresponding change in data and code. Schema may be explicit (database schema, type declarations)
  or implicit (expected shape of data structures or a document).
\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
  database) or less permanent (state in live programming). When scheam evolve, data needs to be
  updated or (in case of transient state) discarded.
\item \emph{Code} -- or program logic, but not including source code that defines schema.
  When schema evolve, code needs to change to consume/produce data of the correct shape.
\end{itemize}

\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their schema, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
schema evolution in future programming systems. We consider two characteristics.

\begin{itemize}
\item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
  eventually adopt all changes. It may not be possible to adopt a change without adopting
  all earlier changes. In the divergence model, a user may choose only
  particular changes they want to adopt, but keep other aspects of their custom design.
\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (schema evolution) can only originate from a particular source. In the decentralized model,
  changes can be done on any of the multiple co-existing variants of a program.
\end{itemize}

\paragraph{Challenge Problems.}
These dimensions, layers, and variants offer a wide range of configurations for challenge problems, but
our aim is not to enumerate all of them. We use them to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain and they also provide and inspiration and a benchmark for
developers of future programming systems.

### 2024-08-23 revisions.tex
Related Work section
revise section 2: Layers and Dimensions of Schema Evolution

@@

Reviewer A

@@

Section 2 has been revised to remove the claim of providing a conceptial framework. Instead we define multiple dimensions of schema change that are used throughout the sequel to figuratively constrast different problems.

Reviewer A
-----------------------------------------------------------------
Not a conceptual framework

>>>>

\documentclass{article}
\usepackage{csquotes}
\usepackage{xcolor}
\begin{document}

@@

\section{Reviewer A}

@@

Section 2 has been revised to remove the claim of providing a conceptual framework. Instead we define a descriptive framework of multiple dimensions and layers of schema change that are used to figuratively contrast the different problems.

\textcolor{red}{I don't know what to do with this:}
\begin{displayquote}
  Just before 2.2 I think here you need to set expectations for correct local transformations, because no general solution exists, only conventions and expectations. Please, add numbers to those expectations, show examples of corner cases where stuff breaks and is an acceptable evil, and cases were stuff breaks on purpose (in some sense, and intended evil) To make an example: Java is of course not truly "cross-platform": if you wanted to put a formal definition to cross platform, the program should work identically in all environments. This would imply that the program should not be able to discover it's running environment. In Java I can make an 'if' at the beginning of the main and throw error on Windows. This makes Java not truly "cross-platform" in a logical sense, but this is often considered desirable.

  In the context of those data transformations, there must be plenty of situations like that, where a change in behavior is not only possible or acceptable, but even desirable. However, we should have some clear formal definition of when/how/why such changes in behavior are introduced.
\end{displayquote}

\section{Reviewer B}

As suggested we have replaced `type' with `schema' throughout the paper. The diagrams still use the letter `T' but we intend to change those to `S' in the camera-ready verison.

As suggested, the Literature Review was converted into a more appropriately named Related Work section.

\section{Reviewer C}

\end{document}

Reviewer A
-----------------------------------------------------------------
I will start from the end. In the conclusion you state: 'we developed a conceptual framework....' I disagree. You informally spoke about those ideas in interesting and tough provoking ways, but this is very far from developing a conceptual framework, and if I had to interpret your work as a conceptual framework I would find it severely lacking. However, I do not need to do so, I think your work can stand on its own as a set of considerations of those problems from a unifying viewpoint. This viewpoint is interesting, but is clearly limited and is basically only applicable to layered systems. With imperative systems I mean system with a set of predefined datatypes that are in some sense 'superior' to the user defined ones, that are just lists/records/unions of those primitive data types. This is not the case for example in lambda calculus, where there are no primitive data types, or in pure OO, where everything truly is an object and you could define string and ints as user defined types using programming patterns not dissimilar to church encoding.

Basically, as soon as 'true abstraction' is into place, most of your considerations stop being applicable.
